---
title: "Causal Inference in R"
subtitle: "Chapter summaries"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
#| echo: false
library(tibble)
library(kableExtra)
```

::: {.callout-important}
## License and attribution

These summaries are a derivative work of *Causal Inference in R* by Malcolm Barrett,
Lucy D'Agostino McGowan, and Travis Gerke, published at
[r-causal.org](https://www.r-causal.org/).

The original work is copyright the authors and licensed under
[CC BY-NC 4.0](https://creativecommons.org/licenses/by-nc/4.0/).
This derivative is released under the same license.

In practice: credit the original authors when you share or adapt this material.
Do not use it commercially. If you are unsure whether your use qualifies as
non-commercial, read the
[CC BY-NC 4.0 deed](https://creativecommons.org/licenses/by-nc/4.0/).
:::

---

The book has one central question: given data you did not design, can you make a
credible causal claim? Not a predictive one. Not a correlational one. A causal
one: what would happen if you *intervened*?

That framing rules a lot out. RCTs are not covered because they sidestep the hard
parts. The whole point of the book is the harder, messier, more common situation:
data collected by someone else, for reasons that may have nothing to do with your
question, where the exposure was not assigned at random and you have to be explicit
about every assumption you are making.

Prerequisites are light: tidyverse fluency, basic R modelling (`lm()`, `glm()`),
and comfort writing functions. tidymodels appears later but no prior exposure is
assumed.

```{r roadmap-table}
#| echo: false
#| tbl-cap: "24 chapters across three parts."

tibble(
  Part = c(
    rep("Part I: Asking Causal Questions", 6),
    rep("Part II: The Design Phase", 3),
    rep("Part III: Estimating Causal Effects", 15)
  ),
  Ch = paste0("Ch ", 1:24),
  Title = c(
    "From casual to causal",
    "The whole game: mosquito nets and malaria",
    "Potential outcomes and counterfactuals",
    "Expressing causal questions as DAGs",
    "Causal inference is not (just) a statistical problem",
    "From question to answer: stratification and outcome models",
    "Preparing data to answer causal questions",
    "Propensity scores",
    "Evaluating your propensity score model",
    "Causal estimands",
    "Fitting the weighted outcome model",
    "Continuous and categorical exposures",
    "G-computation",
    "Interaction",
    "Missingness and measurement",
    "Sensitivity analysis",
    "Causal mediation analysis",
    "Causal inference across time",
    "Causal time-to-event models",
    "Doubly robust models",
    "Machine learning and causal inference",
    "Instrumental variables and friends",
    "Difference-in-difference",
    "Evidence"
  )
) |>
  kbl(
    col.names = c("Part", "Ch", "Title"),
    booktabs = TRUE,
    align = c("l", "c", "l")
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "condensed", "hover"),
    full_width = TRUE
  ) |>
  column_spec(2, bold = TRUE, width = "6%") |>
  collapse_rows(columns = 1, valign = "top")
```

---

## Part I: Asking Causal Questions

Before any method makes sense, you need the framework. What is a causal effect,
exactly? How do you write down your assumptions? Why does data alone never settle
the causation question? Six chapters. They repay slow reading.

---

### Ch 1: From casual to causal

*Association vs. causation · causal questions · counterfactuals*

Most statistical training is about association. This chapter argues that is the
wrong frame for a large class of questions people actually care about. Knowing
that X and Y move together tells you nothing about what happens when you *change* X.

The distinction the book draws between descriptive, predictive, and causal
questions is not pedantic. It has real consequences. A predictive model
optimised for accuracy will often mislead you if you try to use it causally. The
sentence-diagram framework (*does X cause Y, among population Z, compared to
doing W, over horizon T*) is worth internalising early. Every word in that
sentence affects what you are estimating, and dropping one tends to make the
question unanswerable rather than simpler.

---

### Ch 2: The whole game

*End-to-end example · confounding · IPW · DAGs*

A complete causal analysis before any of the theory gets deep. The question is
straightforward: do insecticide-treated mosquito nets reduce malaria risk?

The confounding problem is immediate. Wealthier, healthier households are more
likely to use nets and less likely to get malaria, for reasons unrelated to nets.
Naively comparing users to non-users overstates protection. The chapter walks
through DAG construction, covariate identification, propensity scoring, and
inverse probability weighting, none of it explained in full yet. The point is
to see the whole pipeline once before studying each piece. Reading this chapter
again after finishing the book is genuinely useful.

---

### Ch 3: Potential outcomes and counterfactuals

*Potential outcomes · exchangeability · positivity · consistency · SUTVA · target trial*

The theoretical core. The central problem is that you can only ever observe one
version of what happened. The counterfactual is permanently missing. Causal
inference is, at root, a missing data problem.

The potential outcomes framework defines the individual causal effect as
$Y(1) - Y(0)$. Since only one is observed per person, individual effects are
unidentifiable. The average treatment effect is what we can go after, but only
under three conditions. Exchangeability: no unmeasured confounding, so treatment
groups have the same potential outcomes on average. Positivity: every unit has a
non-zero probability of receiving any exposure level, within all covariate strata.
Consistency (SUTVA): the observed outcome equals the potential outcome for the
received exposure, with no multiple treatment versions and no interference between
units.

Randomisation satisfies all three in the limit. Observational data does not, which
is why the rest of the book exists. The chapter closes with target trial emulation:
write down the RCT you would run if you could, then map each protocol element to
your observed data. It is a more disciplined organising framework than it first
sounds.

---

### Ch 4: Expressing causal questions as DAGs

*DAGs · d-separation · confounding · colliders · mediators · ggdag*

DAGs make assumptions explicit and auditable. A node per variable, a directed edge
per causal relationship, no cycles. The graph does not claim to be the truth. It
commits to your current best guess, which can then be examined and challenged by
people who know the subject matter differently.

D-separation is the graphical rule for reading off conditional independencies,
which translates directly to covariate selection: adjust for backdoor paths, do
not adjust for mediators, and be very careful about colliders.

Colliders are the most counterintuitive part. Conditioning on a common effect of
treatment and outcome (including it in a regression or restricting the sample on
it) opens a spurious association that does not exist in the full population. A
lot of analysts introduce bias while thinking they are tightening up the analysis.
`ggdag` handles visualisation and d-separation queries in R.

---

### Ch 5: Causal inference is not (just) a statistical problem

*Domain knowledge · unmeasured confounding · assumptions · bias*

The hardest chapter to act on, because it tells you something statistical training
actively obscures: no method substitutes for knowing the subject.

The identifiability conditions from Chapter 3, especially exchangeability, cannot
be verified from data. You can check balance, test functional form, run sensitivity
analyses. None of that tells you whether you have adjusted for the right
confounders. That is a question about the world. Two analysts with the same data
but different domain knowledge will reach different conclusions, and both might be
defensible.

More controls is not always safer. Conditioning on a collider is worse than not
adjusting at all. Adjusting for a post-treatment variable introduces bias. Knowing
which variables belong in a model and which do not is inseparable from
understanding the subject matter.

---

### Ch 6: From question to answer

*Stratification · regression adjustment · g-computation · marginal standardisation*

With the framework in place, two approaches to actually estimating an effect.

Stratification compares treatment and control within covariate levels. Transparent,
easy to explain, and it collapses fast. Too many variables or any continuous
covariates and you run out of observations per stratum.

Outcome modelling solves the dimensionality problem. Fit $E[Y \mid X, Z]$, predict
counterfactual outcomes for every unit under each exposure, then average the
difference. This is g-computation, also called marginal standardisation. The key
point is that the ATE is not just the coefficient on $X$ in the outcome model.
That coefficient is only the causal effect under specific functional form
assumptions. The predict-and-average procedure is more general.

---

## Part II: The Design Phase

Chapters 7-9 slow down on a step that is easy to skip: getting the data and
propensity model right before touching the outcome. A propensity model with high
predictive accuracy can still leave substantial confounding, and you will not
catch that from AUC alone.

---

### Ch 7: Preparing data to answer causal questions

*Data preparation · covariate selection · eligibility criteria · temporal ordering*

Before fitting anything: who is included, what the exposure measures, what the
baseline period is, and which covariates are needed for conditional exchangeability.
The DAG does the work here. Covariate selection driven by stepwise procedures or
variable importance is unreliable for causal inference. You need variables that
block backdoor paths, not variables that predict the outcome well. Related, but
different criteria.

Temporal ordering is often overlooked. A covariate measured after the exposure
started cannot be a confounder. Conditioning on it either introduces bias or
blocks a mediation path. Getting this wrong is common and hard to detect from the
data alone.

---

### Ch 8: Propensity scores

*Propensity scores · IPW · tidymodels · overlap · weighting*

The propensity score $e(Z) = P(\text{treatment} \mid Z)$ collapses the covariate
adjustment problem to a single number. Its balancing property means weighting or
stratifying on it achieves the same result as adjusting for all covariates
individually.

The main use is inverse probability weighting: weight each unit by the inverse of
its probability of receiving the treatment it actually received, creating a
pseudo-population where treatment is unrelated to covariates. One thing worth
getting right: a good propensity model is not one with high AUC. It is one that
achieves covariate balance after weighting. Optimising for prediction can actively
hurt balance. Propensity models are fit using `tidymodels`.

---

### Ch 9: Evaluating your propensity score model

*Balance · SMD · love plots · overlap · halfmoon*

Standard predictive diagnostics do not tell you whether a propensity model works
for causal inference. The right diagnostic is standardised mean differences (SMDs)
before and after weighting, which directly show whether covariate distributions
are similar across treatment groups.

Love plots display SMDs for every covariate. The target is below 0.1. Overlap
plots check positivity visually. Extreme weights are a warning sign: some units
were nearly certain to receive or not receive treatment regardless of covariates,
which inflates variance and signals near-positivity violations. Options are to
trim extreme weights, revisit the covariate set, or narrow the target population.
The `halfmoon` package handles these diagnostics.

---

## Part III: Estimating Causal Effects

Chapters 10-24. Standard IPW and g-computation first, then continuous exposures,
time-varying treatments, survival outcomes, effect modification, machine learning,
IV, and DiD. Each chapter builds on the framework from Parts I and II.

---

### Ch 10: Causal estimands

*ATE · ATT · ATC · estimands · target population*

Before fitting a model, decide what you are estimating. The Average Treatment
Effect (ATE) asks what would happen if everyone received treatment versus none.
The Average Treatment Effect on the Treated (ATT) asks what happened to those who
actually received it. These are different quantities and require different
estimators.

ATT is usually more appropriate when the treated group is a specific subgroup and
you want to know whether treatment helped them. ATE fits questions about policies
that would apply broadly. The estimand choice also affects which positivity
violations matter, a detail that compounds through the analysis.

---

### Ch 11: Fitting the weighted outcome model

*IPW · weighted regression · sandwich SE · bootstrap*

With a validated propensity model, fit the outcome model using IPW weights. The
technical complication is standard errors: naive SEs from weighted regression are
wrong because they treat the weights as fixed rather than estimated. Bootstrap SEs
(resample observations, refit both models in each draw) or sandwich SEs fix this.
The chapter is direct about which to prefer and when.

---

### Ch 12: Continuous and categorical exposures

*Continuous treatment · dose-response · generalised propensity score · categorical exposure*

For continuous exposures, the propensity score generalises to a conditional density
$f(X \mid Z)$. Weights from this density stabilise the marginal distribution of
the exposure, and the estimand becomes a dose-response curve rather than a single
average effect. For multi-level categorical exposures the approach extends
naturally; the main decisions are about contrasts and reference categories.

---

### Ch 13: G-computation

*G-computation · marginal standardisation · parametric g-formula*

G-computation is the outcome-modelling alternative to IPW. Fit $E[Y \mid X, Z]$,
predict each unit's counterfactual outcome under treatment and under control,
average the difference. Often more efficient than IPW when the outcome model is
well-specified, and more sensitive to outcome model misspecification when it is
not. That tradeoff motivates the doubly robust methods in Chapter 20.

---

### Ch 14: Interaction

*Effect modification · heterogeneous effects · CATE · additive vs. multiplicative*

The ATE is an average, and averages hide a lot. Treatment effects vary across
subgroups. This is effect modification. The chapter distinguishes statistical
interaction from causal effect modification and covers conditional average
treatment effects (CATEs). Interaction on the additive scale and on the
multiplicative scale are not the same thing, and which scale matters depends on
whether you care about absolute or relative risk differences.

---

### Ch 15: Missingness and measurement

*Missing data · MCAR/MAR/MNAR · multiple imputation · measurement error*

MCAR does not bias estimates but reduces precision. MAR is handled well by
multiple imputation. MNAR requires sensitivity analysis: the missingness depends
on unobserved values, so the mechanism is unidentifiable from data alone.

Measurement error in the exposure attenuates effects toward zero. Measurement
error in confounders is worse: residual confounding remains even after adjustment.
Neither has a clean fix without auxiliary data or strong assumptions, but the
chapter makes the magnitude of bias concrete.

---

### Ch 16: Sensitivity analysis

*Unmeasured confounding · E-value · tipping point · tipr*

The E-value asks how strong an unmeasured confounder would need to be to explain
away the observed effect. Large E-values suggest robustness; small ones signal
fragility. Tipping point analysis finds the unmeasured confounding parameters that
would shift the confidence interval past the null. Both characterise
vulnerability: they tell you how fragile the estimate is, not whether the
lurking confounder actually exists. The `tipr` package implements these in R.

---

### Ch 17: Causal mediation analysis

*Mediation · direct effects · indirect effects · interventional effects*

When you want to know how a treatment works, you need mediation analysis. The
total effect decomposes into direct and indirect effects. The traditional
Baron-Kenny approach does not work for causal inference: it ignores
treatment-mediator interaction and does not handle confounding of the
mediator-outcome path. The chapter uses the interventional effects framework. The
identifying assumptions are stricter than for total effects. Conditional
exchangeability is needed for both the treatment-outcome and mediator-outcome
relationships, with no treatment-induced confounders of the mediator-outcome path.

---

### Ch 18: Causal inference across time

*Longitudinal data · time-varying confounding · marginal structural models*

With time-varying treatments and confounders, standard regression breaks. A
time-varying covariate can simultaneously be a confounder of the current treatment
(adjust for it) and a mediator of past treatment (do not condition on it). Ordinary
regression cannot do both.

Marginal structural models with time-varying IPW handle this. Weights are
constructed from the full treatment history and the outcome model fits the weighted
pseudo-population. The longitudinal g-formula is the parametric alternative. Both
require careful attention to temporal ordering.

---

### Ch 19: Causal time-to-event models

*Survival analysis · hazard ratios · competing risks · RMST · non-collapsibility*

Hazard ratios from Cox models are non-collapsible: even in a randomised trial, the
marginal hazard ratio differs from the conditional one. This is not a data problem.
It is a mathematical property that complicates cross-study comparisons. The
restricted mean survival time (RMST) does not have this property and is more
directly interpretable. Competing risks are also covered: cause-specific and
subdistribution hazards answer different causal questions, and confusing them
changes the conclusion.

---

### Ch 20: Doubly robust models

*AIPW · doubly robust · TMLE · semiparametric efficiency*

Doubly robust estimators combine a propensity model and an outcome model so that
correct specification of either is sufficient for consistency. The augmented IPW
(AIPW) estimator is the main one covered. TMLE directly targets the causal
estimand during estimation. Both are semiparametrically efficient when both models
are correct, and both pair naturally with machine learning nuisance models.

---

### Ch 21: Machine learning and causal inference

*Causal forests · cross-fitting · double machine learning · CATE*

ML nuisance models can fit complex confounding structures without pre-specified
functional form. Used naively, regularisation introduces bias that does not vanish
with sample size. Cross-fitting removes this asymptotically: fit nuisance models
on held-out folds, then rotate. Causal forests directly target heterogeneous
treatment effects without pre-specifying which subgroups to examine. Useful when
the CATEs are the actual question.

---

### Ch 22: Instrumental variables and friends

*IV · LATE · 2SLS · regression discontinuity · natural experiments*

When unmeasured confounding makes exchangeability untenable, IV exploits a
variable that affects treatment but has no direct effect on the outcome. Two things
worth keeping in mind. Valid instruments are hard to find, and the exclusion
restriction is unverifiable from data. IV estimates the LATE (the effect among
compliers), which may not be the population you care about.

Regression discontinuity uses sharp thresholds in treatment assignment to estimate
local causal effects near the boundary. Compelling when the design is clean,
limited in external validity.

---

### Ch 23: Difference-in-differences

*DiD · parallel trends · panel data · staggered adoption · TWFE*

DiD uses a control group's pre-post trajectory as the counterfactual for the
treated group. The identifying assumption is parallel trends: both groups would
have moved together absent treatment.

With staggered adoption, the standard two-way fixed effects regression produces a
weighted average where some weights can be negative even when every individual
effect is positive. The chapter covers heterogeneity-robust estimators designed
for this setting and discusses how pre-treatment periods can be used to assess
(not verify) the parallel trends assumption.

---

### Ch 24: Evidence

*Evidence synthesis · replication · communication · uncertainty*

A single well-executed analysis rarely settles a question. Evidence accumulates
across studies with different designs, populations, and assumptions. The chapter
covers what statistical significance does and does not mean, what replication and
robustness checks add, and how to communicate causal claims honestly, including
the assumptions and their potential failures. Reporting results without making
assumptions visible is not being concise. It is removing the information people
need to evaluate whether your conclusions apply to their context.

---

## Suggested reading paths

```{r reading-paths}
#| echo: false
#| tbl-cap: "Minimum chapter sequences for common goals."

tibble(
  Goal = c(
    "Understand the framework before running anything",
    "Run a complete IPW analysis end-to-end",
    "Handle time-varying treatments",
    "Analyse survival outcomes causally",
    "Estimate heterogeneous treatment effects",
    "Unmeasured confounding is a real concern",
    "Exploit a natural experiment or policy threshold"
  ),
  Path = c(
    "1 → 3 → 4 → 5",
    "1 → 3 → 4 → 7 → 8 → 9 → 10 → 11",
    "1-11 → 18",
    "1-11 → 19",
    "1-11 → 14 → 20 → 21",
    "1-5 → 16 → 22",
    "1-5 → 22 → 23"
  )
) |>
  kbl(
    col.names = c("Goal", "Chapters"),
    booktabs = TRUE,
    align = c("l", "l")
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "condensed", "hover"),
    full_width = TRUE
  ) |>
  column_spec(1, width = "60%") |>
  column_spec(2, bold = TRUE, width = "40%")
```

---

::: {.small}
These summaries are a derivative work of *Causal Inference in R* (Barrett,
D'Agostino McGowan and Gerke, 2026, [r-causal.org](https://www.r-causal.org/)).
The original text is copyright the authors. License:
[CC BY-NC 4.0](https://creativecommons.org/licenses/by-nc/4.0/). Non-commercial
use only.
:::
