```{r}
#| label: setup
#| echo: false

library(ggdag)
library(ggplot2)
library(dplyr)
library(purrr)
library(tibble)
library(patchwork)
library(kableExtra)
library(dagitty)
library(broom)

# Shared theme for all ggdag plots
dag_theme <- theme_dag() %+replace%
  theme(
    legend.position  = "bottom",
    strip.text.x     = element_text(margin = margin(2, 0, 2, 0, "mm")),
    plot.title       = element_text(size = 13, face = "bold", hjust = 0),
    plot.subtitle    = element_text(size = 11, colour = "grey40", hjust = 0)
  )

theme_set(dag_theme)

# Shared ggplot2 theme for data plots
data_theme <- theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    strip.text       = element_text(face = "bold"),
    legend.position  = "bottom"
  )
```

# DAGs

## Overview

DAGs (Directed Acyclic Graphs) let researchers make their causal assumptions explicit and queryable. The chapter covers three things: what DAGs are and how their path types behave statistically, how to build and query them in R using `ggdag`, and practical advice on building defensible DAGs in applied work.

---

## 4.1 Visualizing Causal Assumptions

A DAG is a graph where **nodes** are variables and **edges** are directed arrows representing causal relationships. The direction of an arrow encodes the direction of causation: `x -> y` means `x` causes `y`. Crucially, the graph must be *acyclic*: no variable can be its own ancestor.

DAGs encode a researcher's assumptions about causal structure, not parameter estimates. An arrow says "I believe this cause exists for at least one unit"; it says nothing about the magnitude or functional form of the effect. This non-parametric character distinguishes causal DAGs from structural equation models (SEMs), which require parametric assumptions and estimate the entire graph.

The three elemental path types are **forks**, **chains**, and **colliders**. Every backdoor path decomposes into these structures.

```{r}
#| label: fig-path-types
#| fig-cap: "Three elemental causal path structures. Forks share a common cause; chains pass an effect through a mediator; colliders share a common descendant."

coords <- list(
  x = c(x = 0, y = 2, q = 1),
  y = c(x = 0, y = 0, q = 1)
)

fork     <- dagify(x ~ q, y ~ q, exposure = "x", outcome = "y", coords = coords)
chain    <- dagify(q ~ x, y ~ q, exposure = "x", outcome = "y", coords = coords)
collider <- dagify(q ~ x + y, exposure = "x", outcome = "y", coords = coords)

dag_flows <- map(
  list(fork = fork, chain = chain, collider = collider),
  tidy_dagitty
) |>
  map("data") |>
  list_rbind(names_to = "dag") |>
  mutate(dag = factor(dag, levels = c("fork", "chain", "collider")))

dag_flows |>
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges(edge_width = 1) +
  geom_dag_point(colour = "#2c7bb6", size = 14) +
  geom_dag_text(colour = "white", size = 4) +
  facet_wrap(~dag) +
  expand_plot(
    expand_x = expansion(c(0.2, 0.2)),
    expand_y = expansion(c(0.2, 0.2))
  ) +
  labs(title = "Elemental path structures in a DAG")
```

### Path types and what they imply

```{r}
#| label: tbl-path-types
#| echo: false
#| tbl-cap: "The three path structures, their statistical properties, and when to adjust."

path_tbl <- tibble(
  `Path type`       = c("Fork (confounder)", "Chain (mediator)", "Collider"),
  `Direction of q`  = c("q -> x and q -> y", "x -> q -> y", "x -> q <- y"),
  `Path status`     = c("Open", "Open", "Closed"),
  `x-y correlated?` = c("Yes (spurious)", "Yes (via q)", "No"),
  `Adjust for q?`   = c(
    "Yes, blocks confounding",
    "Depends on question (blocks indirect effect)",
    "No, opens a biasing path"
  )
)

path_tbl |>
  kbl(booktabs = TRUE, align = c("l", "l", "c", "c", "l")) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE,
    font_size         = 13
  ) |>
  column_spec(1, bold = TRUE) |>
  column_spec(5, width = "20em") |>
  row_spec(3, background = "#fff3cd")
```

::: {.callout-note}
## Backdoor paths
Any open, non-causal path from the exposure to the outcome is a **backdoor path**. Forks are the classic backdoor path. Conditioning on a collider can also *open* a previously closed path, creating a backdoor where none existed.
:::

---

### Confounding in practice

```{r}
#| label: fig-confounder-scatter
#| fig-cap: "A fork: q causes both x and y. Unadjusted, x and y appear correlated. Conditioning on q reveals the null relationship."

set.seed(123)
n <- 1000
q_conf <- rbinom(n, size = 1, prob = 0.35)
x_conf <- 2 * q_conf + rnorm(n)
y_conf <- -3 * q_conf + rnorm(n)
confounder_data <- tibble(x = x_conf, y = y_conf, q = as.factor(q_conf))

p_conf1 <- confounder_data |>
  ggplot(aes(x, y)) +
  geom_point(alpha = 0.15, colour = "#636363") +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, colour = "#d7191c", linewidth = 1) +
  facet_wrap(~"Not adjusting for q\n(biased)") +
  data_theme

p_conf2 <- confounder_data |>
  ggplot(aes(x, y, colour = q)) +
  geom_point(alpha = 0.15) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linewidth = 1) +
  scale_colour_manual(values = c("0" = "#2c7bb6", "1" = "#fdae61")) +
  facet_wrap(~"Adjusting for q\n(unbiased)") +
  data_theme

p_conf1 + p_conf2 +
  plot_annotation(
    title    = "Fork: spurious association removed by conditioning on the common cause",
    theme    = theme(plot.title = element_text(size = 13, face = "bold"))
  )
```

### Mediation in practice

```{r}
#| label: fig-mediator-scatter
#| fig-cap: "A chain: adjusting for the mediator q removes the indirect effect and leaves only the direct (null) effect."

set.seed(123)
x_med <- rnorm(n)
lp    <- 2 * x_med + rnorm(n)
q_med <- rbinom(n, size = 1, prob = 1 / (1 + exp(-lp)))
y_med <- 2 * q_med + rnorm(n)
mediator_data <- tibble(x = x_med, y = y_med, q = as.factor(q_med))

p_med1 <- mediator_data |>
  ggplot(aes(x, y)) +
  geom_point(alpha = 0.15, colour = "#636363") +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, colour = "#1a9641", linewidth = 1) +
  facet_wrap(~"Not adjusting for q\n(total effect)") +
  data_theme

p_med2 <- mediator_data |>
  ggplot(aes(x, y, colour = q)) +
  geom_point(alpha = 0.15) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linewidth = 1) +
  scale_colour_manual(values = c("0" = "#2c7bb6", "1" = "#fdae61")) +
  facet_wrap(~"Adjusting for q\n(direct effect)") +
  data_theme

p_med1 + p_med2 +
  plot_annotation(
    title = "Chain: adjustment for mediator isolates the direct effect",
    theme = theme(plot.title = element_text(size = 13, face = "bold"))
  )
```

### Collider bias in practice

```{r}
#| label: fig-collider-scatter
#| fig-cap: "A collider: x and y are independent, but conditioning on q induces a spurious association."

set.seed(123)
x_col   <- rnorm(n)
y_col   <- rnorm(n)
lp_col  <- 2 * x_col + 3 * y_col + rnorm(n)
q_col   <- rbinom(n, size = 1, prob = 1 / (1 + exp(-lp_col)))
collider_data <- tibble(x = x_col, y = y_col, q = as.factor(q_col))

p_col1 <- collider_data |>
  ggplot(aes(x, y)) +
  geom_point(alpha = 0.15, colour = "#636363") +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, colour = "#1a9641", linewidth = 1) +
  facet_wrap(~"Not adjusting for q\n(unbiased)") +
  data_theme

p_col2 <- collider_data |>
  ggplot(aes(x, y, colour = q)) +
  geom_point(alpha = 0.15) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linewidth = 1) +
  scale_colour_manual(values = c("0" = "#2c7bb6", "1" = "#d7191c")) +
  facet_wrap(~"Adjusting for q\n(biased, collider opened)") +
  data_theme

p_col1 + p_col2 +
  plot_annotation(
    title = "Collider: conditioning opens a bias path that wasn't there",
    theme = theme(plot.title = element_text(size = 13, face = "bold"))
  )
```

---

## 4.2 DAGs in R with ggdag

The `ggdag` package wraps `dagitty` with a tidy, ggplot2-compatible interface.

```{r}
#| label: tbl-dagify-args
#| echo: false
#| tbl-cap: "Key arguments to dagify()."

tibble(
  Argument     = c("formulas", "exposure", "outcome", "latent", "coords", "labels"),
  Type         = c("formula...", "character", "character", "character", "list/data.frame", "named character"),
  Purpose      = c(
    "Specify causal relationships (effect ~ cause1 + cause2)",
    "Exposure variable for path queries",
    "Outcome variable for path queries",
    "Mark unmeasured variables; excluded from adjustment sets",
    "Node positions; use time_ordered_coords() for temporal ordering",
    "Human-readable labels for display"
  )
) |>
  kbl(booktabs = TRUE, align = c("l", "l", "l")) |>
  kable_styling(
    bootstrap_options = c("striped", "condensed"),
    full_width        = FALSE,
    font_size         = 13
  ) |>
  column_spec(1, bold = TRUE, monospace = TRUE) |>
  column_spec(2, monospace = TRUE) |>
  column_spec(3, width = "28em")
```

### Podcast exam running example

The running example asks: does listening to a comedy podcast the morning before an exam improve graduate students' test scores?

```{r}
#| label: fig-podcast-dag
#| fig-cap: "Proposed DAG for the podcast exam question. No direct arrow from podcast to exam score is assumed."

podcast_dag <- dagify(
  podcast ~ mood + humor + prepared,
  exam    ~ mood + prepared,
  coords  = time_ordered_coords(
    list(
      c("prepared", "humor", "mood"),
      "podcast",
      "exam"
    )
  ),
  exposure = "podcast",
  outcome  = "exam",
  labels   = c(
    podcast  = "podcast",
    exam     = "exam score",
    mood     = "mood",
    humor    = "humor",
    prepared = "prepared"
  )
)

ggdag(podcast_dag, use_labels = "label", text = FALSE) +
  labs(title = "Podcast exam DAG")
```

### Open paths and adjustment sets

```{r}
#| label: fig-podcast-paths
#| fig-cap: "Two open backdoor paths in podcast_dag: one through mood, one through prepared."

ggdag_paths(podcast_dag, shadow = TRUE, text = FALSE, use_labels = "label") +
  labs(title = "Open paths in podcast_dag")
```

```{r}
#| label: fig-podcast-adjset
#| fig-cap: "Minimal adjustment set for podcast_dag. Both mood and prepared must be adjusted to block all backdoor paths."

ggdag_adjustment_set(podcast_dag, text = FALSE, use_labels = "label") +
  labs(title = "Minimal adjustment set")
```

### Simulated demonstration

```{r}
#| label: fig-sim-correct
#| fig-cap: "Adjusting for the correct set (mood + prepared) recovers the null effect. The unadjusted estimate is spurious."

set.seed(10)
sim_data <- simulate_data(podcast_dag)

unadjusted_model <- lm(exam ~ podcast, sim_data) |>
  tidy(conf.int = TRUE) |>
  filter(term == "podcast") |>
  mutate(formula = "unadjusted")

adjusted_model <- lm(exam ~ podcast + mood + prepared, sim_data) |>
  tidy(conf.int = TRUE) |>
  filter(term == "podcast") |>
  mutate(formula = "mood + prepared")

bind_rows(unadjusted_model, adjusted_model) |>
  ggplot(aes(x = estimate, y = formula, xmin = conf.low, xmax = conf.high)) +
  geom_vline(xintercept = 0, linewidth = 1, colour = "grey75") +
  geom_pointrange(linewidth = 1, size = 0.6, colour = "#2c7bb6") +
  labs(
    x       = "Estimated effect of podcast on exam",
    y       = NULL,
    caption = "True effect = 0",
    title   = "Correct adjustment set recovers the null"
  ) +
  data_theme
```

::: {.callout-warning}
## Wrong DAG, wrong answer
If the DAG is mis-specified, for instance by omitting `mood` as a confounder, then even after adjustment the estimate remains biased. DAG correctness is a prerequisite; the algebra cannot fix a wrong causal story.
:::

---

## 4.3 Structures of Causality

### 4.3.1 Advanced confounding

Backdoor paths need not pass through a single common cause. In a more complex version of the podcast DAG, adding `alertness` (caused by `mood`) and `skills_course` (which frees up time for podcasts and drives preparedness) creates three backdoor paths and four valid minimal adjustment sets.

```{r}
#| label: fig-podcast-dag2
#| fig-cap: "Expanded podcast DAG with alertness and skills_course. Three backdoor paths must be closed."

podcast_dag2 <- dagify(
  podcast      ~ mood + humor + skills_course,
  alertness    ~ mood,
  mood         ~ humor,
  prepared     ~ skills_course,
  exam         ~ alertness + prepared,
  coords       = time_ordered_coords(),
  exposure     = "podcast",
  outcome      = "exam",
  labels       = c(
    podcast      = "podcast",
    exam         = "exam score",
    mood         = "mood",
    alertness    = "alertness",
    skills_course = "college\nskills course",
    humor        = "humor",
    prepared     = "prepared"
  )
)

ggdag(podcast_dag2, use_labels = "label", text = FALSE) +
  labs(title = "Expanded podcast DAG")
```

```{r}
#| label: fig-podcast-dag2-adjsets
#| fig-cap: "Four minimal adjustment sets for podcast_dag2. Each set closes all three backdoor paths."

ggdag_adjustment_set(podcast_dag2, use_labels = "label", text = FALSE) +
  labs(title = "Minimal adjustment sets, expanded DAG")
```

```{r}
#| label: tbl-adjsets
#| echo: false
#| tbl-cap: "Valid minimal adjustment sets for the expanded podcast DAG and considerations for choosing between them."

tibble(
  `Adjustment set`             = c(
    "alertness + prepared",
    "alertness + skills_course",
    "mood + prepared",
    "mood + skills_course"
  ),
  `Backdoor paths closed`      = rep("All three", 4),
  `Practical considerations`   = c(
    "Alertness may be hard to measure accurately",
    "skills_course is objectively verifiable; preferred if well measured",
    "mood + prepared are familiar confounders with moderate measurement quality",
    "skills_course is cleanly measured; mood may vary by self-report quality"
  )
) |>
  kbl(booktabs = TRUE) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE,
    font_size         = 13
  ) |>
  column_spec(1, bold = TRUE, monospace = TRUE) |>
  column_spec(3, width = "24em") |>
  row_spec(2, background = "#d4edda")
```

### 4.3.2 Selection bias and mediation

Selection bias is collider-stratification bias induced by the study design itself, the act of selecting units into the sample. If showing up to the exam (`showed_up`) is caused by both `podcast` and other confounders, inherently conditioning on that variable (by only observing those who showed up) opens a collider path.

```{r}
#| label: fig-podcast-dag3
#| fig-cap: "podcast_dag3: showed_up is both a collider and a mediator. Inherent stratification on this variable limits what causal effect we can estimate."

podcast_dag3 <- dagify(
  podcast    ~ mood + humor + prepared,
  exam       ~ mood + prepared + showed_up,
  showed_up  ~ podcast + mood + prepared,
  coords     = time_ordered_coords(
    list(
      c("prepared", "humor", "mood"),
      "podcast",
      "showed_up",
      "exam"
    )
  ),
  exposure   = "podcast",
  outcome    = "exam",
  labels     = c(
    podcast   = "podcast",
    exam      = "exam score",
    mood      = "mood",
    humor     = "humor",
    prepared  = "prepared",
    showed_up = "showed up"
  )
)

ggdag(podcast_dag3, use_labels = "label", text = FALSE) +
  labs(title = "Selection into sample as a collider-mediator")
```

When the total effect is unrecoverable (indirect path blocked at `showed_up`), the direct effect can sometimes still be estimated.

```{r}
#| label: fig-podcast-dag3-direct
#| fig-cap: "Switching the estimand from total to direct effect yields a valid adjustment set even after conditioning on showed_up."

podcast_dag3 |>
  adjust_for("showed_up") |>
  ggdag_adjustment_set(effect = "direct", text = FALSE, use_labels = "label") +
  labs(title = "Direct effect adjustment set after inherent conditioning on showed_up")
```

#### M-bias and butterfly bias

M-bias occurs when a pre-exposure collider, `m`, is caused by two unmeasured variables `a` and `b` that separately cause the exposure and outcome. The path through `m` is closed by default; conditioning on `m` opens it.

```{r}
#| label: fig-m-bias
#| fig-cap: "M-bias: m is a collider that predates exposure and outcome. The backdoor path x <- a -> m <- b -> y is closed; conditioning on m opens it."

m_bias() |>
  ggdag() +
  labs(title = "M-bias structure")
```

Butterfly (bowtie) bias combines M-bias with confounding: `mood` is both a confounder (must be adjusted) and a collider (adjustment opens a path). When the unmeasured causes are absent from data, there is no clean solution. Empirical guidance: confounding bias tends to be larger than collider bias from M-structures, so adjusting is still the lesser evil.

```{r}
#| label: fig-butterfly
#| fig-cap: "Butterfly bias: mood is simultaneously a confounder and a collider. Controlling for mood blocks confounding but opens the collider path via u1 and u2."

butterfly_bias(x = "podcast", y = "exam", m = "mood", a = "u1", b = "u2") |>
  ggdag(text = FALSE, use_labels = "label") +
  labs(title = "Butterfly (bowtie) bias")
```

### 4.3.3 Instrumental variables and precision variables

```{r}
#| label: tbl-iv-precision
#| echo: false
#| tbl-cap: "Comparison of instrumental variables and precision variables."

tibble(
  Property            = c(
    "Causes exposure?",
    "Causes outcome?",
    "On a backdoor path?",
    "Include in adjustment model?",
    "Effect on point estimate",
    "Effect on variance",
    "Alternative use"
  ),
  `Instrumental variable (e.g., humor)` = c(
    "Yes", "No (by assumption)", "No",
    "Avoid unless suspected confounder",
    "Usually negligible bias if wrongly added",
    "May slightly inflate SE",
    "IV methods (Chapter 22)"
  ),
  `Precision variable (e.g., grader_mood)` = c(
    "No", "Yes", "No",
    "Yes, always include when available",
    "None",
    "Reduces variance, narrower CIs",
    "n/a"
  )
) |>
  kbl(booktabs = TRUE) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE,
    font_size         = 13
  ) |>
  column_spec(1, bold = TRUE) |>
  column_spec(2, width = "16em") |>
  column_spec(3, width = "16em")
```

```{r}
#| label: fig-podcast-dag5
#| fig-cap: "humor is an IV (causes podcast, not exam); grader_mood is a precision variable (causes exam, not podcast)."

podcast_dag5 <- dagify(
  podcast     ~ mood + humor + prepared,
  exam        ~ mood + prepared + grader_mood,
  coords      = time_ordered_coords(
    list(
      c("prepared", "humor", "mood"),
      c("podcast", "grader_mood"),
      "exam"
    )
  ),
  exposure    = "podcast",
  outcome     = "exam",
  labels      = c(
    podcast     = "podcast",
    exam        = "exam score",
    mood        = "student\nmood",
    humor       = "humor",
    prepared    = "prepared",
    grader_mood = "grader\nmood"
  )
)

ggdag(podcast_dag5, use_labels = "label", text = FALSE) +
  labs(title = "IV (humor) and precision variable (grader_mood) in podcast DAG")
```

### 4.3.4 Measurement error and missingness

Separating the *true* value of a variable from its *observed* value in the DAG exposes how measurement bias enters the causal estimate. In recall bias, the outcome influences the exposure measurement error, introducing a non-causal association between observed exposure and outcome.

```{r}
#| label: fig-error-dag
#| fig-cap: "Measurement error DAG: outcome_real causes exposure_error, creating recall bias. The observed exposure is a biased proxy for the true exposure."

error_dag <- dagify(
  exposure_observed ~ exposure_real + exposure_error,
  outcome_observed  ~ outcome_real + outcome_error,
  outcome_real      ~ exposure_real,
  exposure_error    ~ outcome_real,
  labels            = c(
    exposure_real     = "Exposure\n(truth)",
    exposure_error    = "Measurement\nerror (exposure)",
    exposure_observed = "Exposure\n(observed)",
    outcome_real      = "Outcome\n(truth)",
    outcome_error     = "Measurement\nerror (outcome)",
    outcome_observed  = "Outcome\n(observed)"
  ),
  exposure = "exposure_real",
  outcome  = "outcome_real",
  coords   = time_ordered_coords()
)

ggdag(error_dag, text = FALSE, use_labels = "label") +
  labs(title = "Recall bias: outcome influences memory of exposure")
```

---

## 4.4 Recommendations for Building DAGs

### Summary table

```{r}
#| label: tbl-dag-advice
#| echo: false
#| tbl-cap: "Ten recommendations for building defensible DAGs."

tibble(
  `#`             = 1:10,
  Recommendation  = c(
    "Iterate early and often",
    "Consider your question",
    "Order nodes by time",
    "Consider the whole data collection process",
    "Include variables you don't have",
    "Saturate your DAG, then prune",
    "Include instruments and precision variables",
    "Focus on causal structure first, then measurement bias",
    "Pick adjustment sets most likely to succeed",
    "Use robustness checks"
  ),
  Detail          = c(
    "Build the DAG before analysis, ideally before data collection. Share with domain experts and iterate.",
    "The causal structure varies by population and time. What is a confounder in one setting may be irrelevant in another.",
    "Time ordering clarifies assumptions (cause precedes effect) and makes complex DAGs easier to read.",
    "Found data inherits conditioning on the collection mechanism. Understand selection into your sample.",
    "Mark unmeasured variables as latent; ggdag returns only adjustment sets feasible with your data.",
    "Default to including arrows; not-including is the stronger assumption. Prune only implausible arrows.",
    "Neither type changes the adjustment set, but both inform modeling decisions and can flag IV opportunities.",
    "Start with a perfectly-measured DAG; then overlay measurement error as a sensitivity exercise.",
    "When multiple valid sets exist, prefer the one with better-measured, better-modeled variables.",
    "Negative controls, DAG-data consistency checks, and alternate adjustment sets all stress-test your assumptions."
  )
) |>
  kbl(booktabs = TRUE) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = TRUE,
    font_size         = 13
  ) |>
  column_spec(1, width = "2em", bold = TRUE) |>
  column_spec(2, bold = TRUE, width = "14em") |>
  column_spec(3, width = "34em") |>
  row_spec(c(1, 3, 6, 9), background = "#f0f4ff")
```

### Saturate then prune (worked example)

```{r}
#| label: fig-podcast-sat
#| fig-cap: "Fully saturated version of podcast_dag: every variable has arrows to all later variables."

podcast_dag_sat <- dagify(
  podcast  ~ mood + humor + prepared,
  exam     ~ mood + prepared + humor,
  prepared ~ humor,
  mood     ~ humor,
  coords   = time_ordered_coords(
    list("humor", c("prepared", "mood"), "podcast", "exam")
  ),
  exposure = "podcast",
  outcome  = "exam",
  labels   = c(
    podcast  = "podcast",
    exam     = "exam score",
    mood     = "mood",
    humor    = "humor",
    prepared = "prepared"
  )
)

curvatures      <- rep(0, 8)
curvatures[1]   <- 0.25

podcast_dag_sat |>
  tidy_dagitty() |>
  ggplot(aes(x, y, xend = xend, yend = yend, label = label)) +
  geom_dag_point(colour = "#2c7bb6", size = 14) +
  geom_dag_edges_arc(curvature = curvatures) +
  geom_dag_label_repel(aes(label = label), show.legend = FALSE) +
  labs(title = "Saturated podcast DAG")
```

```{r}
#| label: fig-podcast-pruned
#| fig-cap: "Pruned DAG: implausible arrows (humor -> prepared and humor -> exam, since grading is blinded) removed."

podcast_dag_pruned <- dagify(
  podcast  ~ mood + humor + prepared,
  exam     ~ mood + prepared,
  mood     ~ humor,
  coords   = time_ordered_coords(
    list("humor", c("prepared", "mood"), "podcast", "exam")
  ),
  exposure = "podcast",
  outcome  = "exam",
  labels   = c(
    podcast  = "podcast",
    exam     = "exam score",
    mood     = "mood",
    humor    = "humor",
    prepared = "prepared"
  )
)

ggdag(podcast_dag_pruned, text = FALSE, use_labels = "label") +
  labs(title = "Pruned podcast DAG, implausible arrows removed")
```

Both the saturated and pruned DAGs share the same minimal adjustment set (`mood + prepared`), showing that the pruning decision doesn't change what must be adjusted here.

### Feedback loops are feedforward loops

Apparent feedback loops are actually variables co-evolving through time. Expressing this correctly as a time-unrolled DAG avoids the cyclic violation.

```{r}
#| label: fig-feedforward
#| fig-cap: "A/C use and global temperature co-evolve over time. The cyclic mental model unfolds into a feedforward DAG."

dagify(
  global_temp_2000 ~ ac_use_1990 + global_temp_1990,
  ac_use_2000      ~ ac_use_1990 + global_temp_1990,
  global_temp_2010 ~ ac_use_2000 + global_temp_2000,
  ac_use_2010      ~ ac_use_2000 + global_temp_2000,
  global_temp_2020 ~ ac_use_2010 + global_temp_2010,
  ac_use_2020      ~ ac_use_2010 + global_temp_2010,
  coords           = time_ordered_coords(),
  labels           = c(
    ac_use_1990      = "A/C use\n(1990)",
    global_temp_1990 = "Temp\n(1990)",
    ac_use_2000      = "A/C use\n(2000)",
    global_temp_2000 = "Temp\n(2000)",
    ac_use_2010      = "A/C use\n(2010)",
    global_temp_2010 = "Temp\n(2010)",
    ac_use_2020      = "A/C use\n(2020)",
    global_temp_2020 = "Temp\n(2020)"
  )
) |>
  ggdag(text = FALSE, use_labels = "label") +
  labs(title = "Feedback loop unrolled as a time-ordered feedforward DAG")
```

---

## DAG properties in applied health research

The table below summarises empirical properties of DAGs used in 144 applied health research papers (Tennant et al. 2020).

```{r}
#| label: tbl-dag-properties
#| echo: false
#| tbl-cap: "Empirical properties of 144 DAGs in applied health research (Tennant et al. 2020). Most DAGs were only ~46% saturated; fewer than a third reported their estimand."

tibble(
  Characteristic           = c(
    "Number of nodes",
    "Number of arcs",
    "Node-to-arc ratio",
    "Saturation proportion",
    "Fully saturated, Yes",
    "Fully saturated, No",
    "Reported estimand, Yes",
    "Reported estimand, No",
    "Reported adjustment set, Yes",
    "Reported adjustment set, No"
  ),
  `N = 144`                = c(
    "12 (IQR: 9-16)",
    "29 (IQR: 19-42)",
    "2.30 (IQR: 1.75-3.00)",
    "0.46 (IQR: 0.31-0.67)",
    "4 (3%)",
    "140 (97%)",
    "40 (28%)",
    "104 (72%)",
    "80 (56%)",
    "64 (44%)"
  )
) |>
  kbl(booktabs = TRUE) |>
  kable_styling(
    bootstrap_options = c("striped", "condensed"),
    full_width        = FALSE,
    font_size         = 13
  ) |>
  pack_rows("DAG properties", 1, 6) |>
  pack_rows("Reporting", 7, 10) |>
  column_spec(1, width = "20em") |>
  column_spec(2, width = "14em") |>
  row_spec(c(5, 7, 9), bold = TRUE, background = "#fff3cd")
```

---

## Key concepts glossary

```{r}
#| label: tbl-glossary
#| echo: false
#| tbl-cap: "Key terms from Chapter 4."

tibble(
  Term                       = c(
    "DAG", "SCM", "Node / Edge", "Fork",
    "Chain", "Collider", "Open path",
    "Closed path", "Backdoor path",
    "Adjustment set", "Collider-stratification bias",
    "Instrumental variable",
    "Precision variable",
    "Sharp null", "Proxy confounder"
  ),
  Definition = c(
    "Directed Acyclic Graph, a causal diagram with directed arrows and no cycles",
    "Structural Causal Model, the class of non-parametric models that DAGs belong to",
    "Node: a variable. Edge: a directed arrow encoding a causal relationship",
    "q -> x and q -> y: shared common cause; open path; induces confounding",
    "x -> q -> y: effect passes through mediator; open path; blocks causal flow if adjusted",
    "x -> q <- y: shared descendant; closed path; conditioning on it opens a biasing path",
    "A path that transmits statistical association between exposure and outcome",
    "A path that does not transmit association (e.g., at a collider)",
    "Any open, non-causal path from exposure to outcome that must be blocked",
    "A set of variables that, when adjusted, blocks all backdoor paths",
    "Bias introduced by conditioning on a collider (also called selection bias)",
    "A cause of exposure only, not outcome; useful for IV estimation methods",
    "A cause of outcome only, not exposure; reduces variance when included in model",
    "No unit has an individual causal effect; justifies removing an arrow",
    "A measured variable correlated with an unmeasured confounder; reduces but rarely eliminates confounding"
  )
) |>
  kbl(booktabs = TRUE) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = TRUE,
    font_size         = 12
  ) |>
  column_spec(1, bold = TRUE, width = "12em") |>
  column_spec(2, width = "36em")
```

---

::: {.callout-tip}
## Chapter connections
- **Chapter 10** covers estimands in depth (what are you actually trying to estimate?).  
- **Chapter 14** covers interaction and effect modification.  
- **Chapter 15** covers measurement error and missingness.  
- **Chapter 16** covers sensitivity analysis, including robustness checks.  
- **Chapter 17** covers causal mediation analysis.  
- **Chapter 18** covers time-varying confounding and feedforward relationships.  
- **Chapter 22** covers instrumental variable methods.  
:::
