# Chapter3: Potential Outcomes and Counterfactuals

## The Fundamental Problem

A potential outcome is what an outcome variable woult take under a specific exposure(treatment). Before treatmet occurs, every unit has one potential outcome per exposure level, $Y(x)$ for each x. Once an exposure happens, only one potential outcome is realised. The rest are counterfactual ie they exist in the causal model but not in the data.

This is the fundamental problem of causal inference. The individual causal effect $Y(X_{1})$ - $Y(X_{0})$ requirs both potential outcomes simultaneously. You observe exactly one. Every method in the potential outomes framework is an attempt to use observed data to substitute for missing counterfactuals under explicit assumptions that justify the substitution.

## Ice-T, Spike and the counterfactual proxy

The book opens with Ice-T and Spike, two men who grew upin the same Los Angeles gang neighborhoods, ran jewelry heists together, and then diverged. Ice-T was discovered rapping and left the criminal life; Spike continued and received a 35 years to life sentence. The premise is that each serves as the other's counterfactual.

```{r}
#| label: setup
#| include: false
library(tibble)
library(dplyr)
library(stringr)
library(tidyr)
library(kableExtra)
library(ggplot2)
library(purrr)

po_data <- tibble(
  id          = 1:10,
  y_chocolate = c(4, 4, 6, 5, 6, 5, 6, 7, 5, 6),
  y_vanilla   = c(1, 3, 4, 5, 5, 6, 8, 6, 3, 5)
) |>
  mutate(causal_effect = y_chocolate - y_vanilla)

# Colour palette used throughout
pal <- c(
  chocolate  = "#7B3F00",
  vanilla    = "#F5CBA7",
  observed   = "#2c7bb6",
  unobserved = "#d7191c",
  neutral    = "grey50"
)
```

```{r}
#| label: ice-t-spike
#| echo: true

librarian::shelf(tidyverse, kableExtra)

tibble(person = c("Ice-T", "Spike"),
      path = c("Abandoned criminal life", "Continued criminal life"),
      outcome = c("Fame and Fortune", "35 years to life"),
      counterfactual_path = c("Did one more heist", "Abandoned Criminal Life"),
      missing_outcome = c("35 years to life", "Fame and fortune"))  |> 
      kbl(
      align = "lllll",
      col.names = c("Person", "Observed path", 
      "Observed outcome", "Counterfactual path", "Missing outcome")
      ) |>
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
      full_width = FALSE,
      position = "left") |> 
      column_spec(1, bold = TRUE) |> 
      column_spec(4:5, color = "grey40", italic = TRUE) |> 
      add_header_above(c(" " = 3, "Counterfactual (unobservable)" = 2))
```

The premise is appealig but does not hold perfectly. Musical talent is a plausible cause of Ice-T's outcome independent of his decision to quit, which makes Spike a reasonable proxy for what Ice-T would have faced had he continued, but not necessarily a symmetric counterfactual the other way round. This is the same asymmetry problem in every observational analysis. The chapter uses the story to motivate why we work with averaged effects over population rather than individual counterfactuals, and why randomization or adjustment is needed to construct valid proxies.

::: {.callout-note}
## Potential Outcomes vs Counterfactuals

A **Counterfactual** is a potential outcome that was not realised, it's counter to the fact of what happened. Potential outcomes are defined before the exposure, agnostic to what actually occurs. In practice, the book treats the two terms as interchangeable; once exposure has happened, you're always looking at one realised outcome and at least one counter to the fact

:::

## The ice cream simulation

### The oracle table: all potential outcomes known
Ten individuals each have a chocolate happiness score and a vanila happiness score. This "Oracle" view, both potential outcomes visible simultaneously, is never available in practice, it exists here only to define what we are trying to estimate.

```{r}
#| label: tbl-oracle-table
#| echo: true
#| caption: "Oracle potential outcomes table. In real data, only one column per row is ever observable"
library(tibble)
library(kableExtra)

# Save the tibble to a variable so we can reference it in row_spec
po_display <- po_data |>
  rename(
    ID                       = id,
    `Y(chocolate)`           = y_chocolate,
    `Y(vanilla)`             = y_vanilla,
    `Y(choc) - Y(van)`       = causal_effect
  )

po_display |>
  kbl(align = "cccc") |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE,
    position          = "left"
  ) |>
  add_header_above(c(" " = 1, "Potential outcomes" = 2, "Individual causal effect" = 1)) |>
  row_spec(0, bold = TRUE) |>
  row_spec(which(po_display$`Y(choc) - Y(van)` < 0), color = "#c0392b") |>
  row_spec(nrow(po_display), bold = TRUE) |>
  footnote(
    general = paste0(
      "True ATE = ", round(mean(po_data$y_chocolate), 2),
      " - ", round(mean(po_data$y_vanilla), 2),
      " = ", round(mean(po_data$causal_effect), 2),
      ". Rows in red have negative individual causal effects (vanilla > chocolate for that person)."
    ),
    general_title = ""
  )

```

## What we actually observe under randomization

Assign each participant randomly (50/50) to one flavor. Now only the assigned potential outcome is observed; the counterfactual is missing. Individual causal effects become unknowable.

```{r}
#| label: random-assignment
set.seed(11)
observed_rand <- po_data |>
  mutate(
    exposure         = if_else(rbinom(n(), 1, 0.5) == 1, "chocolate", "vanilla"),
    observed_outcome = if_else(exposure == "chocolate", y_chocolate, y_vanilla),
    y_chocolate_disp = if_else(exposure == "chocolate", as.character(y_chocolate), "--"),
    y_vanilla_disp   = if_else(exposure == "vanilla",   as.character(y_vanilla),   "--")
  )

observed_rand |>
  select(
    ID                 = id,
    Exposure           = exposure,
    `Y(chocolate)`     = y_chocolate_disp,
    `Y(vanilla)`       = y_vanilla_disp,
    `Individual effect`= observed_outcome
  ) |>
  mutate(`Individual effect` = "--") |>
  kbl(align = "cclcc") |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE,
    position          = "left"
  ) |>
  add_header_above(c(" " = 2, "Potential outcomes (one missing)" = 2, " " = 1)) |>
  row_spec(0, bold = TRUE) |>
  column_spec(2, bold = TRUE) |>
  column_spec(5, color = "grey50", italic = TRUE) |>
  footnote(
    general = paste0(
      "Observed ATE estimate = ",
      round(mean(observed_rand$observed_outcome[observed_rand$exposure == "chocolate"]), 2),
      " - ",
      round(mean(observed_rand$observed_outcome[observed_rand$exposure == "vanilla"]), 2),
      " = ",
      round(
        mean(observed_rand$observed_outcome[observed_rand$exposure == "chocolate"]) -
        mean(observed_rand$observed_outcome[observed_rand$exposure == "vanilla"]), 2
      ),
      ". Close to true ATE (0.8) due to randomization."
    ),
    general_title = ""
  )
```

## Why Randomization Recovers ATE & self-selection destroys it

```{r}
set.seed(113)
pref_obs <- po_data |>
  mutate(
    prefer_choc      = y_chocolate > y_vanilla,
    exposure         = case_when(
      prefer_choc  ~ if_else(rbinom(n(), 1, 0.8) == 1, "chocolate", "vanilla"),
      !prefer_choc ~ if_else(rbinom(n(), 1, 0.8) == 1, "vanilla",   "chocolate")
    ),
    observed_outcome = if_else(exposure == "chocolate", y_chocolate, y_vanilla)
  )

rand_means <- observed_rand |>
  summarise(mean_h = mean(observed_outcome), .by = exposure) |>
  mutate(mechanism = "Randomization")

pref_means <- pref_obs |>
  summarise(mean_h = mean(observed_outcome), .by = exposure) |>
  mutate(mechanism = "Self-selection (80% prefer)")

plot_means <- bind_rows(rand_means, pref_means) |>
  mutate(
    mechanism = factor(mechanism, levels = c("Randomization", "Self-selection (80% prefer)")),
    exposure  = str_to_title(exposure)
  )

ggplot(plot_means, aes(x = exposure, y = mean_h, colour = exposure, shape = exposure)) +
  geom_hline(yintercept = c(mean(po_data$y_chocolate), mean(po_data$y_vanilla)),
             linetype = "dashed", linewidth = 0.5, colour = "grey60") +
  geom_point(size = 5) +
  geom_segment(
    aes(xend = exposure, yend = 4),
    linewidth = 0.7, show.legend = FALSE
  ) +
  annotate("text", x = 0.6, y = mean(po_data$y_chocolate) + 0.1,
           label = "True avg Y(choc) = 5.4", hjust = 0, size = 3, colour = "grey40") +
  annotate("text", x = 0.6, y = mean(po_data$y_vanilla) - 0.15,
           label = "True avg Y(van) = 4.6", hjust = 0, size = 3, colour = "grey40") +
  facet_wrap(~mechanism) +
  scale_colour_manual(values = c(Chocolate = pal["chocolate"], Vanilla = "#c8a870"),
                      name = NULL) +
  scale_shape_manual(values = c(Chocolate = 19, Vanilla = 17), name = NULL) +
  scale_y_continuous(limits = c(3.5, 7), breaks = seq(3.5, 7, by = 0.5)) +
  labs(x = NULL, y = "Observed mean happiness") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    strip.text         = element_text(face = "bold", size = 12),
    legend.position    = "bottom"
  )

```
```{r}
ggplot(po_data, aes(x = causal_effect)) +
  geom_dotplot(
    binwidth   = 0.5,
    fill       = pal["observed"],
    colour     = "white",
    stackratio = 1.1
  ) +
  geom_vline(xintercept = mean(po_data$causal_effect),
             linetype = "dashed", linewidth = 0.8, colour = pal["unobserved"]) +
  annotate("text",
           x     = mean(po_data$causal_effect) + 0.12,
           y     = 0.9,
           label = paste0("ATE = ", round(mean(po_data$causal_effect), 1)),
           hjust = 0, size = 3.8, colour = pal["unobserved"]) +
  scale_x_continuous(breaks = -3:4, limits = c(-3, 4)) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(x = "Individual causal effect: Y(chocolate) - Y(vanilla)") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
```

## Causal Assumptions

For observed data to stand in missing counterfactuals, three **identifiablity conditions** have to hold. The book cals these the *unconfoundedness* assumptions, theyre required by the IPW, regression adjustment, g-computation, and most other methods covered. Miss one and you're not estimating a causal effect anymore, whatever the statistics say.

```{r}
#| label: tbl-assumptions-overview
#| tbl-cap: "The three core identifiability conditions for unconfoundedness methods."

tibble::tibble(
  Assumption     = c("Exchangeability", "Positivity", "Consistency"),
  `Also called`  = c(
    "No confounding; ignorability; unconfoundedness",
    "Probabilistic assumption; overlap",
    "SUTVA (Stable Unit Treatment Value Assumption)"
  ),
  `Formal notation` = c(
    "\\(Y(x) \\perp\\!\\!\\!\\perp X\\)",
    "\\(P(X = x) > 0\\) for all \\(x\\)",
    "\\(Y_{\\text{obs}} = X \\cdot Y(1) + (1-X) \\cdot Y(0)\\)"
  ),
  `Plain language` = c(
    "Exposure assignment is independent of potential outcomes",
    "Every unit has non-zero probability of each exposure level",
    "Observed outcome = potential outcome under received exposure"
  )
) |>
  kbl(
    escape   = FALSE,
    align    = c("l", "l", "c", "l"),
    booktabs = TRUE
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = TRUE
  ) |>
  column_spec(1, bold = TRUE) |>
  column_spec(3, monospace = TRUE)
```

::: {.callout-important}
### The apples-to-apples principle

All three assumptions exist to solve the same problem; making sure the comparison groups are exchangeable, close enough that each can reasonably stand in for the other's counterfactual. English speakers say "comparing apples to oranges"; other languages have their own versions:

- Cheese and Chalk (UK)
- Grandmothers and toads (Serbian)
- Potatoes and sweet potatoes (Latin American Spanish)

The diagnostic question: do the groups differ in ways that are tied to both the exposure and the outcome? If yes, the comparison is broken before any statistics happen
:::

### Exchangeability

